---
title: "Project 1 Code"
author: "Seth Bitney, Sarah Hall, Nathan Koh, Hannah Richardson"
date: "2024-10-27"
output: html_document
---

```{r setup, include=TRUE}
library(tidyverse)
data <- read.csv('diamonds4.csv')
```

## Individual Variable Code


### Color

```{r}
#reorder bar chart so that it goes from near-colorless (J,I,H,G) to colorless (F,E,D)
data$color <-factor(data$color, levels =c("J","I","H","G","F","E","D"))
#data$color <-factor(data$color, levels =c("D","E","F","G","H","I","J"))

#plot the bar chart
ggplot(data, aes(x=color,y=price))+
  geom_bar(stat="identity", fill = "tomato3")+
  scale_y_continuous(labels = scales::comma)+
  theme(axis.text.x = element_text(),
        plot.title=element_text(hjust = 0.5))+
  labs(x="Color",y="Price", title = "Distribution of Diamond Price by Color Grade")

```
 *Fig 1. Distribution of Diamonds *\


To provide additional context regarding the color variable, we can create a new dataframe which shows the proportion of the number of diamonds in each color group, the average carat, average price, and how frequent it is within the dataset.  

```{r}
Data2<-data %>% 
  group_by(color) %>% 
  summarize(avg.prices=mean(price),avg.carat=mean(carat),total = n()) 

grouped <-table(data$color)
totals<-sum(grouped)

color_percent<-(grouped / totals)*100

new_df<-data.frame(Data2,color_percent)

new_df<-new_df[,c(1,2,3,4,6)] #removing duplicated rows for clarity
new_df
```




```{r}
ggplot(new_df, aes(x=color,y=avg.prices))+
  geom_bar(stat="identity", fill="darkslategrey", col="white")+
  scale_y_continuous(labels = scales::comma)+
  theme(axis.text.x = element_text(),
        plot.title=element_text(hjust = 0.5))+
  labs(x="Color",y="Average Price", title = "Distribution of Average Diamond Price by Color")
```
*Fig. 2: Dataframe showing average price across each diamond color grade*\

```{r}
ggplot(new_df, aes(x=color,y=Freq))+
  geom_bar(stat="identity",fill="tomato3")+
  theme(axis.text.x = element_text(angle=0),
        plot.title=element_text(hjust = 0.5))+
  labs(x="Color",y="Frequency", title = "Frequency of Diamonds across the Color Grade Scale")
```


*Note: Due to the extreme outliers across all color grades, this made it difficult to see the boxplots themselves. Due to this, two boxplots were created (Fig.4 and Fig.5) for clarity. The first boxplot contains the outliers, while the second boxplot is rescaled so that the plots themselves are visible.*

```{r }
ggplot(data,aes(x=color,y=price,fill=color))+
  geom_boxplot()+
  scale_y_continuous(labels = scales::comma)+
  theme(axis.text.x = element_text(),plot.title=element_text(hjust = 0.5))+
  labs(x="Color",y="Price", title = "Boxplot of Diamond Color Against Price, with Outliers")
```
*Fig.4: Color-Price Boxplot with Outlier Indicators*

```{r}
ggplot(data,aes(x=color,y=price,fill=color))+
  geom_boxplot(outlier.color=NA)+ylim(0,12500)+
  theme(axis.text.x = element_text(),plot.title=element_text(hjust = 0.5))+
  labs(x="Color",y="Price",title="Color to Diamond Price Distribution, without Outliers")
  
```

### Clarity

```{r}
Clarity_Table <- table(data$clarity)
round(prop.table(Clarity_Table), 4)

data$clarity <-factor(data$clarity, levels =c("SI2","SI1","VS2","VS1","VVS2","VVS1","IF", "FL"))
```


```{r}
ggplot(data, aes(x=clarity, y=price))+
  geom_bar(stat="identity", fill="darkslategrey")+
  scale_y_continuous(labels = scales::comma)+
  theme(axis.text.x = element_text(angle = 0),
        plot.title = element_text(hjust = 0.5))+
  labs(x="Clarity", y="Price",title="Distribution of Clarity vs Cumulative Price")
```


```{r}
# Alt plot by mean
ggplot(data, aes(x=clarity, y=price))+
  scale_y_continuous(labels = scales::comma)+
  stat_summary(fun = "mean", geom = "bar",fill="darkslategrey")+
  theme(axis.text.x = element_text(angle = 0),
        plot.title = element_text(hjust = 0.5))+
  labs(x="Clarity", y="Price",title="Distribution of Clarity vs Average Price")

group_mean<- aggregate(x= data$price,
                      # Specify group indicator
                      by = list(data$clarity),      
                      # Specify function (i.e. mean)
                      FUN = mean)
print(group_mean)
```


```{r}
ggplot(data, aes(x=clarity, y=price))+
  geom_violin(fill="tomato3")+
  scale_y_continuous(labels = scales::comma)+
  theme(axis.text.x = element_text(angle = 0),
        plot.title = element_text(hjust = 0.5))+
  labs(x="Clarity", y="Price", title="Distribution of Prices by Clarity")
```



```{r}
ggplot(data, aes(x=clarity,y=price))+
  geom_point(alpha=0.5,col="darkslategrey",size=3)+
  scale_y_continuous(labels = scales::comma)+
  labs(x="Clarity", y="Price",title="Scatterplot of Blue Nile Diamond Clarity Against Price")+
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
data$clarity_modified <- gsub('[0-9.]', '', data$clarity)
data

data$clarity_modified <-factor(data$clarity_modified, levels =c("SI","VS","VVS","IF", "FL"))
# Sarah's plot with modified distribution
ggplot(data, aes(x=clarity_modified, y=price))+
  scale_y_continuous(labels = scales::comma)+
  stat_summary(fun = "mean", geom = "bar",fill="darkslategrey")+
  theme(axis.text.x = element_text(angle = 0),
        plot.title = element_text(hjust = 0.5))+
  labs(x="Clarity", y="Price",title="Distribution of Clarity Grouped vs Average Price")

group_mean<- aggregate(x= data$price,
                       # Specify group indicator
                       by = list(data$clarity_modified),      
                       # Specify function (i.e. mean)
                       FUN = mean)
print(group_mean)



#Diamond Clarity by Color
ggplot(data, aes(x=clarity,fill=color))+
  geom_bar(stat="count", width=0.7)+
  theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Diamond Clarity by Color",x="Diamond Clarity", y="Number of Diamonds")
```

### Carat 

```{r}
meancarat<-mean(data$carat)
#Mean carat size is .813
mediancarat<-median(data$carat)
#Median carat size is .52

#Checking the spread of just carat
hist(data$carat,xlab="Carat Weight",ylab="# of Diamonds",main="Distribution of Blue Nile Diamond Carat Size", col='darkslategrey',border='white')


```

```{r}
#Checking the relationship between price and carat

ggplot(data, aes(x=carat, y=price)) +
  geom_point(size=2.5, col="tomato3",alpha=0.4)+
  scale_y_continuous(labels = scales::comma)+
  labs(x="Carat Weight", y="Price",title="Scatterplot of Carat Weight Against Price of Blue Nile Diamonds")+
  theme(plot.title = element_text(hjust = 0.5))
```



```{r}
ggplot(data, aes(x=cut, y=price, size=carat)) +
  geom_point(col="darkslategrey",alpha=0.4)+
  scale_y_continuous(labels = scales::comma)+
  labs(x="Carat Weight", y="Price",title="Scatterplot of Cut and Carat Against Price of Blue Nile Diamonds")+
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
#Get ratio of carat weight per dollar
pricepercarat<-(data$price/data$carat)
data<-data.frame(data,pricepercarat)

data$caratratio <-round(data$pricepercarat,3)

data$cut<-factor(data$cut, levels=c("Good","Very Good","Ideal","Astor Ideal"))

ggplot(data,aes(x=cut,y=pricepercarat))+
  geom_boxplot()+
  scale_y_continuous(labels = scales::comma)+
  theme(axis.text.x = element_text(),plot.title=element_text(hjust = 0.5))+
  labs(x="Cut",y="Price Per Carat", title = "Diamond Cut Against Price per Carat, with Outliers")

ggplot(data, aes(x=carat, y=pricepercarat)) +
  geom_point(size=2.5, col="darkslategrey",alpha=0.3)+
  scale_y_continuous(labels = scales::comma)+
  labs(x="Carat Weight", y="Price",title="Scatterplot of Carat Weight Against Price of Blue Nile Diamonds")+
  theme(plot.title = element_text(hjust = 0.5))

```

### Cut 

```{r}
ggplot(data = data, mapping = aes(x=cut,y=price))+
  scale_y_continuous(labels = scales::comma)+
  geom_point(size=2.5, col="darkslategrey",alpha=0.3)+
   theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  geom_smooth(method = lm, se = FALSE)+
  labs(x="Cut", y = "Price", title = "Diamond Price by Cut Quality")

ggplot(data, aes(x=cut, y=price))+
  geom_violin()+
  labs(x="cut", y="price", title="Diamond cut vs price")


median_data <- aggregate(data$price,         # Median by group
          list(data$cut),
          median)
median_data

mean_data <- aggregate(data$price,         # mean by group
          list(data$cut),
          mean)
mean_data 


ggplot(median_data, aes(x=Group.1, y=x))+
  geom_bar(fill= 'tomato3', fill="white", stat="identity")+ 
  theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  labs(x="Cut", y="Median Price", title="Median Price by Cut")

ggplot(mean_data, aes(x=Group.1, y=x))+
  geom_bar(fill= 'darkslategrey', fill="white", stat="identity")+ 
  theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  labs(x="Cut", y="Average Price", title="Average Price by Cut")
```

```{r}
# Stacked
ggplot(data, aes(x=color,fill=cut))+
  geom_bar(stat="count", width=0.7)+
    theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Diamond Color by Cut",x="Diamond Color", y="Number of Diamonds")
```


```{r}
ggplot(data, aes(x=color,fill=clarity))+
  geom_bar(stat="count", width=0.7)+
    theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Diamond Color by Clarity",x="Diamond Color", y="Number of Diamonds")
```


```{r}
# Stacked
ggplot(data, aes(x=clarity,fill=cut))+
  geom_bar(stat="count", width=0.7)+
    theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  labs(title = "Diamond Clarity by Cut",x="Diamond Clarity", y="Number of Diamonds")
```


### Price


```{r}
meanprice<-mean(data$price)
#Mean price is $7,056.74
medianprice<-median(data$price)
#Median price is $1,463.50
```


```{r}
ggplot(data, aes(x=carat, y=price))+
  geom_point(col="tomato3",alpha=0.6,size=2)+ #scatterplot
      theme(axis.text.x = element_text(), 
        plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(labels = scales::comma)+
  geom_smooth(method = "lm", se = FALSE,col="darkslategrey")+ #adds linear line 
  labs(x="Carat", y="Price", title="Effect of Diamond Carat against Price")

```
 

```{r}
caratregr <- lm(price~carat, data=data)
summary(caratregr)
```


```{r}
par(mfrow = c(2, 2))
plot(caratregr)
```

```{r}
library(MASS) ##to use boxcox function
MASS::boxcox(caratregr, lambda = seq(-0.5, 0.5, 1/10))

MASS::boxcox(caratregr, lambda = seq(0.2, 0.4, 1/10))
```
 
```{r}
#transform y 
data$newy<-log(data$price)

#regress new y against x
result <- lm(newy~carat, data = data)

#create a new residual plot 
par(mfrow=c(2,2))
plot(result)
```


```{r}
ggplot2::ggplot(data, aes(x=carat,y=newy))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
      theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(hjust = 0.5))+
  labs(x="Carat", y="Log of Price",
  title="Log of Price against Carat")

```


```{r}
data$caratstar<-log(data$carat)

ggplot(data, aes(x=caratstar, y=newy))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
      theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(hjust = 0.5))+
  labs(x="Log of Carat", y="Log of Price", title="Effect of Regressed Diamond Carat against Regressed Price")

```


```{r}
final_result <- lm(newy~caratstar, data = data)

#create a new residual plot 
par(mfrow=c(2,2))
plot(final_result)
```


```{r}
cor(data$caratstar, data$newy)
```


```{r}
summary(final_result)
```

```{r}
anova.tab <- anova(final_result)
anova.tab
```
 

```{r}
qf(0.95,1,1212)
```

